---
title: "Assignment3"
author: "Rohit Koonireddy"
date: "2023-12-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**Exercise A**
```{r message=FALSE, warning=FALSE}
#import relevant libraries
library(mlogit)
library(dfidx)
library(AER)
library(nnet)

# load the data
load("DataTransport.RData")
str(Data)

#check na columns
na_count_by_column <- colSums(is.na(Data))
print(na_count_by_column)

#transport data is in wide format
transport_data <- Data
library(dplyr)

transport_data <- transport_data %>%
  mutate(choice = ifelse(choice == 1, "yes", "no"))

transport_data <- transport_data %>%
  select(id, mode, choice, everything())

transport_data <- distinct(transport_data)
head(transport_data,20)
```
```{r}
#take unique values
transport_data_unique <- transport_data %>%
  distinct(id, mode, .keep_all = TRUE)

str(transport_data_unique)
```

```{r}
#transform data to long
transport <- dfidx(transport_data_unique, 
                idx = list("id", "mode"), 
                choice = "choice", # not really needed here
                levels = c("Helicopter","WaterTaxi","Ferry","Hovercraft"))

transport <- na.omit(transport)
head(transport)
with(transport_data_unique, prop.table(table(mode[choice == "yes"])))
```
```{r}
colnames(transport_data_unique)
```


A1) Estimate a multinomial logit model about the choice of transportation mode. You can include as 
many controls as you want, but you need to include at least the cost and the risk attributes. Show the 
coefficient results. NOTE: The risk variable is constant for each option. It is thus possible to have 
collinearity problems if you include individual characteristics in the model. Feel free to use only the 
attributes/individual characteristics that allow the model to work.
```{r}
# Fit a multinomial logit model
mlogit_model <- mlogit(choice ~ cost + risk + noise + crowdness+ convloc+clientele| 0, data=transport)

# Summarize the results
summary(mlogit_model)
```

A2) What is the marginal effect of a change in the cost of the ferry on the probability of taking the 
water taxi? Explain what that number means. NOTE: as we saw in class, the marginal effect formula 
depends on ð‘ƒð‘–ð‘— and therefore on the exact value of the choice attributes/individual characteristics. 
For this exercise, use for each choice the average of each attribute and the average of the individual 
characteristics (if you include any of them).

1 unit chage in the cost of ferry will increase the probability of taking water taxi by 11.67% as shown.
```{r}
#different way to run the marginal effect
effects(mlogit_model, covariate = "cost", type = "ar")
```

A3) Calculate the WTP to reduce passenger/trip death. Explain how you calculate this value and its 
meaning. HINT: look at the slides and the papers discussed in class.

WTP increase by 0.25 for reduced risk.
```{r}
# Extract coefficients
coef_risk <- coef(mlogit_model)["risk"]
coef_cost <- coef(mlogit_model)["cost"]

WTP_risk_cost <- coef_cost/coef_risk
WTP_risk_cost
```
A4) Estimate the same model in question A1 by using a nested logit model with 2 nests. Justify your 
choice of the branches/nests. Show the resulting coefficients and explain the advantages and 
disadvantages of this model compared to the multinomial logit estimated in question A1.
```{r}
nested_model <- mlogit::mlogit(choice ~ cost + risk + noise + crowdness+ convloc+clientele | 0 | mode, transport)

cat("Nested Logit Model:", "\n")
summary(nested_model)
```

BONUS QUESTION: Answer question A5 for extra points on your assignment grade
A5) Estimate the same model in question A1 by using a random coefficients logit. Explain which
coefficient(s) you impose as random and why you think is a good assumption. Show the coefficient 
estimates and explain the advantages and disadvantages of this model compared to the multinomial 
logit estimated in question A1.

I think risk perception and the noise level can be relative and can be judged by various factors, I chose these 2 as my random variables.
```{r}
random_model <- mlogit::mlogit(choice ~ cost + risk + noise + crowdness+ convloc+clientele  | 0, transport, rpar = c( risk = "n", noise="n"))

cat("Random Coefficients Logit Model:", "\n")
summary(random_model)
```


**Exercise B**
In the file DataAIDS.RData you will find data from Chang and Serletis (2014). The data, originally taken 
from the 1997-2009 Survey of Household Spending of Canada, contains information on expenditure in 
the transportation sector of 2218 Canadian households made by single individuals. The transportation 
expenditure is divided between gasoline (good 1), local public transport (good 2) and intercity 
transport like buses (good 3). The data contains the year (year), the total transportation expenditure 
(y), the share of transportation expenditure spent for each transport option (s1, s2, s3), the normalized 
2002 prices of the good based on the consumer price index (p1, p2, p3).1
Use the micEconAids package to run an Almost Ideal Demand System for the transportation choices of 
the sample and answer the following questions:

B1) Calculate the own- and cross-elasticities (choose and comment only one between 
Hicksian/compensated or Walrasian/Marshallian/uncompensated throughout the whole exercise) of 
the three transportation choices using an AIDS model and without approximating the price index (i.e. 
using the iterated linear least squares procedure). Give a brief interpretation of the results underlining 
the key findings.
```{r message=FALSE, warning=FALSE}
# Load the micEconAids package
library(micEconAids)

# Load the data
load("DataAIDS.RData")

# Explore the structure of the data
str(Data)

priceNames <- c("p1", "p2", "p3")
shareNames <- c("s1", "s2", "s3")

# Run the AIDS model using the iterated linear least squares procedure (IL)
est_result_IL <- aidsEst(priceNames, shareNames, "y", data = Data, method = "IL")

# Display summary of the results
summary(est_result_IL)

# Calculate own- and cross-elasticities
elas_IL <- elas(est_result_IL)
print("B1) Own- and Cross-Elasticities (Hicksian/Compensated):")
print(elas_IL)
```

B2) Calculate the own- and cross-elasticities of the three transportation choices using an AIDS model 
and using the Stone price index as linear approximation of the price index. Compare the results 
obtained with and without approximating the price index. What do you notice? Do you believe that in 
this case the results using the Stone price index linearization are acceptable compared to those 
obtained with model using the ILLS in question B1?
```{r}
est_result_LA_S <- aidsEst(priceNames, shareNames, "y",
                          data = Data,
                          method = "LA",  # Linear Approximation
                          priceIndex = "S")  # Use Stone index

# Display summary of the results
summary(est_result_LA_S)

```
The LA-S model, using the Stone price index, generally yields more elastic responses compared to the IL model.
The choice of the linear approximation method can impact the magnitude and direction of elasticities.

1 That means that 1 is the value of 2002 and represents the reference value. For instance, if the price for 2005 
is 1.07 it means that the price increased by 7% compared to 2002.
